---
title: 'Project: '
author: "Hossam Osama Ahmed - 21010451"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Setup
```{r}
# Install required packages if not already installed
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("SNPRelate", "GENESIS", "GWASTools", "qqman", "GeneNet", "gplots"))
install.packages(c("readxl", "writexl", "ggplot2", "dplyr", "tidyr"))
```

```{r}


# Load libraries
library(SNPRelate)
library(GENESIS)
library(GWASTools)
library(qqman)
library(GeneNet)
library(gplots)
library(readxl)
library(writexl)
library(ggplot2)
library(dplyr)
library(tidyr)

```


## Task 1: Compute Kinship using SNPRelate and GENESIS

### 1a. Compute kinship matrix

```{r}

# Convert PLINK files to GDS format
snpgdsPED2GDS("Qatari156_filtered_pruned.ped", 
               "Qatari156_filtered_pruned.map", 
               "Qatari156_filtered_pruned.gds", 
               family=TRUE)

# Open the GDS file
genofile <- snpgdsOpen("Qatari156_filtered_pruned.gds")

# Calculate IBD coefficients
ibd <- snpgdsIBDKING(genofile, num.thread=2)

# Close the file
snpgdsClose(genofile)

# Extract kinship matrix
kinship <- ibd$kinship
rownames(kinship) <- ibd$sample.id
colnames(kinship) <- ibd$sample.id

# Save kinship matrix
saveRDS(kinship, "kinship_matrix.rds")
```

### 1b. Report individuals with kinship > 0.1

```{r}
# Find pairs with kinship > 0.1
high_kinship_pairs <- which(kinship > 0.1, arr.ind = TRUE)
high_kinship_ind <- unique(rownames(high_kinship_pairs))

# Count and report
num_high_kinship <- length(high_kinship_ind)
cat("Number of individuals with kinship > 0.1:", num_high_kinship, "\n")
```

## Task 2: Compute mQTLs with Mixed Models

### 2a-c. mQTL analysis with mixed models
```{r}
# Load required libraries
library(GENESIS)
library(gdsfmt)
library(SNPRelate)
library(openxlsx)
library(dplyr)

# ----------------------------------------
# 1. Set Up File Paths (MODIFY THESE)
# ----------------------------------------

metab_file <- "qatari_metabolites_2025.csv"
output_file <- "mQTL_results.xlsx"

# ----------------------------------------
# 2. Data Preparation with Error Checking
# ----------------------------------------

snpgdsClose(gds)


# Convert PLINK to GDS format
gds_file <- "Qatari156_filtered_pruned.gds"
if (!file.exists(gds_file)) {
  snpgdsPED2GDS(ped_file, map_file, gds_file)
}

# Open GDS file with error handling
genofile <- tryCatch({
  snpgdsOpen(gds_file)
}, error = function(e) {
  stop(paste("Failed to open GDS file:", e$message))
})

# Load metabolite data
metabolites <- read.csv(metab_file, row.names = 1)

# Verify sample matching
if (!all(rownames(metabolites) %in% read.gdsn(index.gdsn(genofile, "sample.id"))) {
  warning("Sample IDs in metabolite data don't match genetic data!")
}

# ----------------------------------------
# 3. Kinship and PCA Calculation
# ----------------------------------------

# Calculate kinship matrix
ibd <- snpgdsIBDKING(genofile, num.thread = 2)
kinship <- ibd$kinship
rownames(kinship) <- colnames(kinship) <- ibd$sample.id

# Calculate first 3 PCs
pca <- snpgdsPCA(genofile, num.thread = 2)
pcs <- data.frame(
  scanID = pca$sample.id,
  PC1 = pca$eigenvect[,1],
  PC2 = pca$eigenvect[,2],
  PC3 = pca$eigenvect[,3]
)

# ----------------------------------------
# 4. mQTL Analysis with Progress Tracking
# ----------------------------------------

# Initialize storage
all_results <- list()
heritability_results <- list()
processed <- 0
errors <- 0

for (metab in colnames(metabolites)) {
  processed <- processed + 1
  cat(sprintf("\nProcessing %d/%d: %s", processed, ncol(metabolites), metab)
  
  # Prepare phenotype data
  pheno <- data.frame(
    scanID = rownames(metabolites),
    metabolite = metabolites[,metab]
  )
  pheno <- merge(pheno, pcs, by = "scanID")
  
  # Skip if too many missing values
  if (sum(is.na(pheno$metabolite)) > 0.1 * nrow(pheno)) {
    cat(" - Skipped (too many NAs)")
    errors <- errors + 1
    next
  }
  
  # Fit null model
  nullmod <- tryCatch({
    fitNullModel(
      pheno,
      outcome = "metabolite",
      covars = c("PC1", "PC2", "PC3"),
      cov.mat = kinship,
      verbose = FALSE
    )
  }, error = function(e) {
    cat(" - Null model failed:", e$message)
    errors <- errors + 1
    NULL
  })
  
  if (is.null(nullmod)) next
  
  # Run association testing
  assoc <- tryCatch({
    assocTestSingle(
      genoData = genofile,
      null.model = nullmod,
      test = "Score",
      verbose = FALSE
    )
  }, error = function(e) {
    cat(" - Association failed:", e$message)
    errors <- errors + 1
    NULL
  })
  
  if (is.null(assoc)) next
  
  # Store results
  assoc$metabolite <- metab
  all_results[[metab]] <- assoc
  
  # Calculate heritability
  h2 <- varCompCI(nullmod, prop = TRUE)
  heritability_results[[metab]] <- data.frame(
    metabolite = metab,
    h2 = h2$estimates["V_stand", "estimate"],
    h2_lower = h2$ci["V_stand", "lower"],
    h2_upper = h2$ci["V_stand", "upper"],
    stringsAsFactors = FALSE
  )
}

# ----------------------------------------
# 5. Process and Save Results
# ----------------------------------------

# Combine all associations
all_assoc <- bind_rows(all_results)

# Filter significant hits (p < 0.0001)
sig_assoc <- all_assoc %>% 
  filter(pval < 0.0001) %>%
  transmute(
    metabolite,
    SNP = variant.id,
    chr = as.character(chr),
    position = pos,
    effect_allele = allele.index,
    sample_size = n.obs,
    beta = Score,
    se = Score.SE,
    p_value = pval
  )

# Combine heritability results
heritability_df <- bind_rows(heritability_results)

# Save to Excel
wb <- createWorkbook()
addWorksheet(wb, "Significant_mQTLs")
writeData(wb, "Significant_mQTLs", sig_assoc)
addWorksheet(wb, "Heritability")
writeData(wb, "Heritability", heritability_df)
saveWorkbook(wb, output_file, overwrite = TRUE)

# ----------------------------------------
# 6. Clean Up and Summary
# ----------------------------------------

# Close GDS file
snpgdsClose(genofile)

# Print summary
cat("\n\n=== ANALYSIS COMPLETE ===\n")
cat("Metabolites processed:", processed, "\n")
cat("Successful analyses:", processed - errors, "\n")
cat("Errors encountered:", errors, "\n")
cat("Significant associations found:", nrow(sig_assoc), "\n")
cat("Results saved to:", output_file, "\n")

# Return top results if any
if (nrow(sig_assoc) > 0) {
  cat("\nTop significant associations:\n")
  print(head(sig_assoc, 5))
} else {
  cat("\nNo significant associations at p < 0.0001 threshold\n")
}

```


```{r}
# ------------------------
# TASK 2: Fixed mQTL Analysis using y/X/cov.mat GENESIS method
# ------------------------

# 1. Load libraries
library(SNPRelate)
library(GENESIS)
library(openxlsx)
library(Matrix)
library(SeqVarTools)
library(GWASTools)


# 2. Open genotype GDS
gds_file <- "Qatari156_filtered_pruned.gds"

gds <- snpgdsOpen(gds_file)

# 3. Compute PCA
pca <- snpgdsPCA(gds, eigen.cnt = 3, num.thread = 2)
pcs <- data.frame(
  sample.id = pca$sample.id,
  PC1 = pca$eigenvect[, 1],
  PC2 = pca$eigenvect[, 2],
  PC3 = pca$eigenvect[, 3],
  row.names = pca$sample.id
)

# 4. Load kinship matrix and make it positive definite
# Replace this with your real kinship matrix loading step
# Assuming `kinship` was already computed from Task 1:
# You must have something like this:
# kinship <- readRDS("kinship_matrix.rds") or however you saved it

# Ensure kinship matrix is symmetric and positive definite
kinship <- nearPD(kinship, corr = FALSE, keepDiag = TRUE)$mat
kinship <- as.matrix(kinship)
rownames(kinship) <- colnames(kinship)

# 5. Load metabolite data
metab <- read.csv("qatari_metabolites_2025.csv", row.names = 1)
rownames(metab) <- gsub("\\*", "", rownames(metab))

# 6. Align samples
common_samples <- Reduce(intersect, list(
  rownames(metab),
  rownames(kinship),
  rownames(pcs)
))
metab <- metab[common_samples, ]
kinship <- kinship[common_samples, common_samples]
pcs <- pcs[common_samples, ]

# 7. Prepare output containers
results_all <- list()
heritabilities <- data.frame(
  Metabolite = character(),
  Heritability = numeric(),
  CI_lower = numeric(),
  CI_upper = numeric(),
  stringsAsFactors = FALSE
)

# 8. Loop through each metabolite
for (met in colnames(metab)) {
  cat("Processing", met, "...\n")
  
  # Outcome
  y <- metab[common_samples, met]
  
  # Covariates matrix
  X <- model.matrix(~ PC1 + PC2 + PC3, data = pcs[common_samples, ])

  # Create a data frame for the null model
  dat <- data.frame(y = y, pcs[common_samples, ])

  # Fit null model using GENESIS syntax
  nullmod <- tryCatch({
    GENESIS::fitNullModel(
      dat,
      outcome = "y",
      covars = c("PC1", "PC2", "PC3"),
      cov.mat = kinship,
      family = "gaussian"
    )
  }, error = function(e) {
    cat("❌ fitNullModel failed for", met, ":", e$message, "\n")
    return(NULL)
  })

  if (!is.null(nullmod)) {
    # Try to get heritability
    var_comp <- tryCatch(varCompCI(nullmod), error = function(e) NULL)

    if (!is.null(var_comp) && "Kinship" %in% rownames(var_comp$varComp)) {
      h2 <- var_comp$varComp["Kinship", "estimate"]
      
      if ("Kinship" %in% rownames(var_comp$CI)) {
        h2_CI <- var_comp$CI["Kinship", ]

        heritabilities <- rbind(heritabilities, data.frame(
          Metabolite = met,
          Heritability = h2,
          CI_lower = h2_CI["lower"],
          CI_upper = h2_CI["upper"]
        ))
      } else {
        cat("⚠️ CI for Kinship not available for", met, "\n")
        heritabilities <- rbind(heritabilities, data.frame(
          Metabolite = met,
          Heritability = h2,
          CI_lower = NA,
          CI_upper = NA
        ))
      }
    } else {
      cat("⚠️ Kinship component missing or varCompCI failed for", met, "\n")
      heritabilities <- rbind(heritabilities, data.frame(
        Metabolite = met,
        Heritability = NA,
        CI_lower = NA,
        CI_upper = NA
      ))
    }

    # Association testing
    assoc <- tryCatch({
      assocTestSingle(
        null.model = nullmod,
        GdsGenotypeReader(gds_file)
      )
    }, error = function(e) {
      cat("❌ assocTestSingle failed for", met, ":", e$message, "\n")
      return(NULL)
    })

    # Keep significant
    if (!is.null(assoc)) {
      sig <- assoc[assoc$Score.pval < 1e-4, ]
      if (nrow(sig) > 0) {
        sig$Metabolite <- met
        results_all[[met]] <- sig
      }
    }
  }
}

# 9. Export significant results
if (length(results_all) > 0) {
  combined <- do.call(rbind, results_all)

  final_results <- data.frame(
    SNP = combined$variant.id,
    Chr = combined$chr,
    Position = combined$pos,
    Metabolite = combined$Metabolite,
    Effect_Allele = combined$allele.1,
    Beta = combined$Est,
    SE = combined$SE,
    P_value = combined$Score.pval,
    N = combined$n.obs,
    stringsAsFactors = FALSE
  )

  write.xlsx(final_results, "significant_mQTLs.xlsx", row.names = FALSE)
} else {
  message("No significant SNP-metabolite associations found (p < 0.0001).")
}

# 10. Export heritabilities
write.xlsx(heritabilities, "heritabilities.xlsx", row.names = FALSE)

# 11. Cleanup
snpgdsClose(gds)
cat("✅ Task 2 completed successfully.\n")


```


```{r}
# Load phenotype (metabolite) and covariate data
pheno <- read.csv("qatari_metabolites_2025.csv")
covars <- read.csv("covariates.csv")  # i dont have this or pc1 pr pc2 

# Add first 3 PCs to covariates
covars <- cbind(covars, pc1, pc2, pc3)

# Create genotype object
geno <- GdsGenotypeReader("Qatari156_filtered_pruned.gds")
genoData <- GenotypeData(geno)

# Fit null model (accounts for kinship and covariates)
null_model <- fitNullModel(pheno, 
                          outcome = "metabolite_level",
                          covars = covars,
                          cov.mat = kinship,
                          family = gaussian())

# Perform association testing
assoc_results <- assocTestSingle(genoData, 
                                null.model = null_model,
                                test = "Score")

# Save results
write.csv(assoc_results, "mQTL_results.csv")

```
### 2d. Report significant associations (p < 0.0001)


```{r}
# Filter significant results
sig_results <- assoc_results[assoc_results$p.value < 0.0001, ]

# Select required columns
report_table <- sig_results[, c("snpID", "chr", "pos", "beta", "SE", 
                               "effect.allele", "n.obs", "p.value")]

# Save to Excel
write_xlsx(report_table, "significant_mQTLs.xlsx")
```


### 2e. Report heritability

```{r}
# Calculate variance components
var_comp <- varCompCI(null_model)

# Extract heritability
h2 <- var_comp$h2
cat("Estimated heritability:", h2, "\n")

```

## Task 3: Inflation Factor Calculation
```{r}
# Calculate lambda (inflation factor) for each metabolite
lambda_values <- sapply(unique(assoc_results$metabolite), function(metab) {
    pvals <- assoc_results$p.value[assoc_results$metabolite == metab]
    median(qchisq(pvals, df=1, lower.tail=FALSE)) / qchisq(0.5, 1)
})

# Calculate average lambda
avg_lambda <- mean(lambda_values)

# Report results
lambda_df <- data.frame(Metabolite = names(lambda_values), 
                       Lambda = lambda_values)
write.csv(lambda_df, "inflation_factors.csv")
cat("Average inflation factor:", avg_lambda, "\n")
```

## Task 4: Manhattan Plot 
```{r}
# Prepare data for Manhattan plot
manhattan_data <- assoc_results %>%
    select(chr, pos, p.value, metabolite) %>%
    rename(CHR = chr, BP = pos, P = p.value, SNP = metabolite)

# Create Manhattan plot
png("manhattan_plot.png", width=1000, height=600)
manhattan(manhattan_data, 
          main = "mQTL Manhattan Plot",
          suggestiveline = -log10(1e-3), 
          genomewideline = -log10(5e-8),
          col = c("blue4", "orange3"))
dev.off()
```

## Task 5: Metabolic Networks
###

```{r}
# Correct metabolite levels
corrected_metabs <- residuals(null_model)

# Save corrected metabolites
write.csv(corrected_metabs, "corrected_metabolites.csv")
```

### 5b. Find significant metabolite pairs with GeneNet

```{r}
# Compute partial correlations
pcor <- ggm.estimate.pcor(corrected_metabs)

# Test for significance
network <- network.test(pcor, corrected_metabs)

# Extract significant pairs
sig_pairs <- extract.network(network, cutoff.ggm = 0.95)

# Save results
write.csv(sig_pairs, "metabolite_network_pairs.csv")
```

### 5c-d. Visualize network in Cytoscape

```{r}
# Prepare node attributes
nodes <- data.frame(id = colnames(corrected_metabs),
                    name = colnames(corrected_metabs))

# Prepare edge list
edges <- sig_pairs %>%
    select(node1, node2, edge.weights) %>%
    rename(source = node1, target = node2, weight = edge.weights)

# Save for Cytoscape
write.csv(nodes, "network_nodes.csv", row.names=FALSE)
write.csv(edges, "network_edges.csv", row.names=FALSE)
```

## Task 6: Annotate Significant SNPs

```{r}
# Get top 20 unique SNPs
top_snps <- assoc_results %>%
    arrange(p.value) %>%
    distinct(snpID, .keep_all = TRUE) %>%
    head(20)

# Prepare ANNOVAR input
annovar_input <- top_snps %>%
    select(chr, pos, pos, ref, alt) %>%
    mutate(chr = paste0("chr", chr))

write.table(annovar_input, "annovar_input.txt", 
            sep="\t", col.names=FALSE, row.names=FALSE, quote=FALSE)

```

### annotate_variation.pl annovar_input.txt humandb/ -buildver hg19 -out myanno -dbtype ensGene

## Task 7: Regional Plots using SNIPA

```{r}
# Get top 5 unique SNPs
top5_snps <- assoc_results %>%
    arrange(p.value) %>%
    distinct(snpID, .keep_all = TRUE) %>%
    head(5)

# Prepare SNIPA input
snipa_input <- top5_snps %>%
    select(chr, pos, snpID)

write.table(snipa_input, "snipa_input.txt", sep="\t", row.names=FALSE)

```






